### 2018-02-12 submission/manuscript.Rmd
update with bringing the lead to the fore
update abstract, intro, and methods

**Methods.** DNA from Fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach. First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communities. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase. Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.

@@

Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced biases, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017]. Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results.

Currently, preservation media used, storage conditions, and DNA extraction kits chosen are the most commonly studied biases. The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g. Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/). Within the literature, DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017]. However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015]. Although these are important sources of bias they are not the only sources that should be critically examined. The type of DNA polymerase chosen could have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that the number of OTUs generated were not easily removed using the authors chosen bioinformatic pipeline regardless of DNA polymerase used [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

@@

***Human and Mock Samples:***
A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples have been published previously [@erin_seekatz_2016]. Clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used in this study and is made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance  (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR Protocol:***
The five different high fidelity (HiFi) DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified. For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples = 100). Although, the mock communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130). No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

@@

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people. Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000 sequence depth. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). Based on these observations we wanted to next analyze potential reasons for these differences. First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing. For the community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).

>>>>

**Methods.** DNA from fecal samples (n = 4) were extracted using a single PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities (technical replicates n = 4) consisting of previously isolated whole genomes of 8 different bacteria were also amplified using the same PCR amplification approach. First, the number of OTUs (Operational Taxonomic Units) was examined for both fecal samples and mock communities. Next, Bray-Curtis index, the error rate, sequence error prevalence, and chimera prevalence were assessed based on cycle number and HiFi DNA polymerase. Finally, the chimera prevalence correlation with number of OTUs was assessed for both cycle number and HiFi DNA polymerase dependent differences.

@@

Recently there has been an increasing focus on standardizing methodological approaches in microbiota research [@review_Kim_2017; @review_Hugerth_2017]. In particular, investigating ways that 16S rRNA gene surveys can be made more reproducible has been a predominant driver of this standardization push [@contamination_Salter2014; @taq_Gohl_2016; @storage_Lauber_2010; @preservation_Song_2016]. Although 16S rRNA gene sequencing has been much maligned for introduced biases, many of these same considerations also affect metagenomic sequencing [@metagenomcis_bias_Nayfach_2016; @metagenomcis_bias_Costea_2017]. Between the two approaches similar bias considerations include preservation media, storage conditions, DNA extraction kit, PCR, and sequence library preparation. Thus, for these overlapping considerations, biases identified for 16S rRNA gene sequencing will also likely influence metagenomic sequencing results. This current study specifically addresses the PCR contribution of bias to this overall workflow.

Currently, preservation media used, storage conditions, and DNA extraction kits chosen are the most commonly studied biases. The study of these specific biases has become so large, aggregating them all together has become a difficult task and some researchers provide resources to actively track new findings (e.g. Microbiome Digest - https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/). Within the literature, DNA extraction kits have consistently been shown to add bias to downstream analysis [@contamination_Salter2014; @metagenomcis_bias_Costea_2017]. However, the current literature on preservation media and storage conditions has been more mixed, with some studies showing biases while others do not [@preservation_Song_2016; @preservation_Luo_2016; @preservation_Dominianni_2014; @storage_Lauber_2010; @storage_Bassis_2017; @preservation_Sinha_2015]. Although these are important sources of bias they are not the only sources that should be critically examined. The type of DNA polymerase chosen could also have a wide ranging affect on downstream results due to error rates and chimeras that may not be easily resolved using bioinformatic approaches.

A recent study in *Nature Biotechnology* showed that there were clear differences between normal and high fidelity (HiFi) DNA polymerase and that you could reduce error and chimera generation by optimizing the PCR protocol [@taq_Gohl_2016]. Another important component of this study was that the number of Operational Taxonomic Units (OTUs) generated were not easily removed using the authors chosen bioinformatic pipeline, regardless of DNA polymerase used [@taq_Gohl_2016]. Although it is probably not surprising that normal DNA polymerase performed worse than HiFi DNA polymerase, it is natural to extend this line of inquiry and ask whether different HiFi DNA polymerase contribute different biases to downstream sequencing results. There is some reason to think that this may be the case since many of these HiFi DNA polymerase come from different families (e.g. *Taq* belongs to the family A polymerases) and may intrinsically have different error rates that cannot be completely removed with modifications [@polymerase_Ishino_2014]. In this study we critically examine if any of five different HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

@@

***Human and Mock Samples:***
A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples have been published previously [@erin_seekatz_2016]. Other than confirmation that none of these individuals had a *Clostridium difficle* infection clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used in this study and is made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance  (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR Protocol:***
The five different high fidelity (HiFi) DNA polymerases that were tested included AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified. For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each HiFi DNA polymerase (total samples = 100). The mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130). No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

@@

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data transformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people. Bray-Curtis distance matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000 sequence depth. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and Kruskal-Wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curtis indices were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). For these community based measures, the fecal samples were analyzed at 4 different sub-sampling sequence depths (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000). Next, based on these observations we wanted to analyze potential reasons for these differences. First, analysis of general sequence error rate, number of sequences with an error, and base substitution were assessed in the mock community for each DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing.

### 2018-02-12 submission/manuscript.tex
update with bringing the lead to the fore
update abstract, intro, and methods

\textbf{Methods.} DNA from Fecal samples (n = 4) were extracted using a
single PowerMag DNA extraction kit with a 10 minute bead beating step
and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same PCR
amplification approach. First, the number of OTUs (Operational Taxonomic
Units) was examined for both fecal samples and mock communities. Next,
Bray-Curtis index, the error rate, sequence error prevalence, and
chimera prevalence were assessed based on cycle number and HiFi DNA
polymerase. Finally, the chimera prevalence correlation with number of
OTUs was assessed for both cycle number and HiFi DNA polymerase
dependent differences.

@@

Recently there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more reproducible has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced biases, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include preservation media, storage conditions, DNA
extraction kit, PCR, and sequence library preparation. Thus, for these
overlapping considerations biases identified for 16S rRNA gene
sequencing will also likely influence metagenomic sequencing results.

Currently, preservation media used, storage conditions, and DNA
extraction kits chosen are the most commonly studied biases. The study
of these specific biases has become so large, aggregating them all
together has become a difficult task and some researchers provide
resources to actively track new findings (e.g.~Microbiome Digest -
\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
Within the literature, DNA extraction kits have consistently been shown
to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
2017). However, the current literature on preservation media and storage
conditions has been more mixed, with some studies showing biases while
others do not (Lauber et al., 2010; Dominianni et al., 2014; Sinha et
al., 2015; Song et al., 2016; Luo et al., 2016; Bassis et al., 2017).
Although these are important sources of bias they are not the only
sources that should be critically examined. The type of DNA polymerase
chosen could have a wide ranging affect on downstream results due to
error rates and chimeras that may not be easily resolved using
bioinformatic approaches.

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that the number of OTUs generated were not easily removed
using the authors chosen bioinformatic pipeline regardless of DNA
polymerase used (Gohl et al., 2016). Although it is probably not
surprising that normal DNA polymerase performed worse than HiFi DNA
polymerase, it is natural to extend this line of inquiry and ask whether
different HiFi DNA polymerase contribute different biases to downstream
sequencing results. There is some reason to think that this may be the
case since many of these HiFi DNA polymerase come from different
families (e.g. \emph{Taq} belongs to the family A polymerases) and may
intrinsically have different error rates that cannot be completely
removed with modifications (Ishino \& Ishino, 2014). In this study we
critically examine if any of five different HiFi DNA polymerases
introduce significant biases into 16S rRNA gene surveys, if this is a
cycle dependent phenomenon, and whether they can be removed using a
standard bioinformatic pipeline.

@@

\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
obtained from 4 individuals who were part of the Enterics Research
Investigational Network (ERIN). The processing and storage of these
samples have been published previously (Seekatz et al., 2016). Clinical
data and other types of meta data were not utilized or accessed for this
study. All samples were extracted using the MOBIO\textsuperscript{TM}
PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The
ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
CA, USA) was used in this study and is made up of \emph{Pseudomonas
aeruginosa}, \emph{Escherichia coli}, \emph{Salmonella enterica},
\emph{Lactobacillus fermentum}, \emph{Enterococcus faecalis},
\emph{Staphylococcus aureus}, \emph{Listeria monocytogenes}, and
\emph{Bacillus subtilis} at equal genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, the protocol
previously published by Gohl and colleagues was used (Gohl et al.,
2016). For Phusion, the company defined conditions were used except for
extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased
by 5 up to 35 cycles with amplicons used at each 5-step increase for
sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30,
and 35 cycles for each Taq (total samples = 100). Although, the mock
communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10
replicates for 30 cycles for all HiFi DNA polymerase (total samples =
130). No mock community sample had enough PCR product at 15 cycles for
adequate 16S rRNA gene sequencing.

@@

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation between people. Bray-Curtis matrices were generated
using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000
sequence depth. The distance matrix data was analyzed using PERMANOVA
with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
Kruskal-Wallis tests within R. For both error and chimera analysis,
samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
Where applicable correction for multiple comparison utilized the
Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curtis indices were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). Based on these observations we wanted to next analyze
potential reasons for these differences. First, analysis of general
sequence error rate, number of sequences with an error, and base
substitution were assessed in the mock community for each DNA
polymerase. After assessing these errors, the total number of chimeras
was determined after sequence processing. For the community based
measures, the fecal samples were analyzed at 4 different sub-sampling
sequence depths (1000, 5000, 10000, and 15000) while the mock community
samples were analysed at 3 levels (1000, 5000, 10000).

>>>>

\textbf{Methods.} DNA from fecal samples (n = 4) were extracted using a
single PowerMag DNA extraction kit with a 10 minute bead beating step
and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Mock communities
(technical replicates n = 4) consisting of previously isolated whole
genomes of 8 different bacteria were also amplified using the same PCR
amplification approach. First, the number of OTUs (Operational Taxonomic
Units) was examined for both fecal samples and mock communities. Next,
Bray-Curtis index, the error rate, sequence error prevalence, and
chimera prevalence were assessed based on cycle number and HiFi DNA
polymerase. Finally, the chimera prevalence correlation with number of
OTUs was assessed for both cycle number and HiFi DNA polymerase
dependent differences.

@@

Recently there has been an increasing focus on standardizing
methodological approaches in microbiota research (Kim et al., 2017;
Hugerth \& Andersson, 2017). In particular, investigating ways that 16S
rRNA gene surveys can be made more reproducible has been a predominant
driver of this standardization push (Lauber et al., 2010; Salter et al.,
2014; Song et al., 2016; Gohl et al., 2016). Although 16S rRNA gene
sequencing has been much maligned for introduced biases, many of these
same considerations also affect metagenomic sequencing (Nayfach \&
Pollard, 2016; Costea et al., 2017). Between the two approaches similar
bias considerations include preservation media, storage conditions, DNA
extraction kit, PCR, and sequence library preparation. Thus, for these
overlapping considerations, biases identified for 16S rRNA gene
sequencing will also likely influence metagenomic sequencing results.
This current study specifically addresses the PCR contribution of bias
to this overall workflow.

Currently, preservation media used, storage conditions, and DNA
extraction kits chosen are the most commonly studied biases. The study
of these specific biases has become so large, aggregating them all
together has become a difficult task and some researchers provide
resources to actively track new findings (e.g.~Microbiome Digest -
\url{https://microbiomedigest.com/microbiome-papers-collection/microbiome-techniques/sample-storage/}).
Within the literature, DNA extraction kits have consistently been shown
to add bias to downstream analysis (Salter et al., 2014; Costea et al.,
2017). However, the current literature on preservation media and storage
conditions has been more mixed, with some studies showing biases while
others do not (Lauber et al., 2010; Dominianni et al., 2014; Sinha et
al., 2015; Song et al., 2016; Luo et al., 2016; Bassis et al., 2017).
Although these are important sources of bias they are not the only
sources that should be critically examined. The type of DNA polymerase
chosen could also have a wide ranging affect on downstream results due
to error rates and chimeras that may not be easily resolved using
bioinformatic approaches.

A recent study in \emph{Nature Biotechnology} showed that there were
clear differences between normal and high fidelity (HiFi) DNA polymerase
and that you could reduce error and chimera generation by optimizing the
PCR protocol (Gohl et al., 2016). Another important component of this
study was that the number of Operational Taxonomic Units (OTUs)
generated were not easily removed using the authors chosen bioinformatic
pipeline, regardless of DNA polymerase used (Gohl et al., 2016).
Although it is probably not surprising that normal DNA polymerase
performed worse than HiFi DNA polymerase, it is natural to extend this
line of inquiry and ask whether different HiFi DNA polymerase contribute
different biases to downstream sequencing results. There is some reason
to think that this may be the case since many of these HiFi DNA
polymerase come from different families (e.g. \emph{Taq} belongs to the
family A polymerases) and may intrinsically have different error rates
that cannot be completely removed with modifications (Ishino \& Ishino,
2014). In this study we critically examine if any of five different HiFi
DNA polymerases introduce significant biases into 16S rRNA gene surveys,
if this is a cycle dependent phenomenon, and whether they can be removed
using a standard bioinformatic pipeline.

@@

\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
obtained from 4 individuals who were part of the Enterics Research
Investigational Network (ERIN). The processing and storage of these
samples have been published previously (Seekatz et al., 2016). Other
than confirmation that none of these individuals had a \emph{Clostridium
difficle} infection clinical data and other types of meta data were not
utilized or accessed for this study. All samples were extracted using
the MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used in this study and is
made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
DNA polymerases that were tested included AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, the protocol
previously published by Gohl and colleagues was used (Gohl et al.,
2016). For Phusion, the company defined conditions were used except for
extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased
by 5 up to 35 cycles with amplicons used at each 5-step increase for
sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30,
and 35 cycles for each HiFi DNA polymerase (total samples = 100). The
mock communities had 4 replicates at 15, 20, 25, and 35 cycles and 10
replicates for 30 cycles for all HiFi DNA polymerase (total samples =
130). No mock community sample had enough PCR product at 15 cycles for
adequate 16S rRNA gene sequencing.

@@

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation between people. Bray-Curtis distance matrices were
generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
15000 sequence depth. The distance matrix data was analyzed using
PERMANOVA with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
Kruskal-Wallis tests within R. For both error and chimera analysis,
samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
Where applicable correction for multiple comparison utilized the
Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curtis indices were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). For these community based measures, the fecal samples
were analyzed at 4 different sub-sampling sequence depths (1000, 5000,
10000, and 15000) while the mock community samples were analysed at 3
levels (1000, 5000, 10000). Next, based on these observations we wanted
to analyze potential reasons for these differences. First, analysis of
general sequence error rate, number of sequences with an error, and base
substitution were assessed in the mock community for each DNA
polymerase. After assessing these errors, the total number of chimeras
was determined after sequence processing.
